Prediction of the 2020 Presidential Election

### 2020-11-02 outputs/paper/paper.Rmd
added to model + plots for results and discussion

library(tidyverse)
library(tinytex)
library(readr)
library(here)
library(dplyr)
library(sjPlot)
```

@@

To accurately predict the winner of the United States 2020 presidential election is difficult, as many different factors are involved.
.....

@@

As previously mentioned, our data is from two different datasets; the Democracy Fund + UCLA Nationscape Wave 50 dataset, and the 2018 1-year American Community Survey (ACS) dataset. They each contain both qualitative and quantitative data. In regard to the Nationscape Wave 50 dataset, a few of its strengths are the variety in variables considered, and the quality of the survey responses. With respect to its weaknesses, it is likely that the data contains sampling error and/or coverage error. One of the various strengths of the  ACS dataset is its ability to be customized with pre-specified (or pre-selected?) variables and sample size before download. Although there is a wide range of variables to choose from, being able to select only those which we needed made the exploratory data analysis process much more manageable than it could have otherwise been (with all possible variables). Furthermore, since sample size can be pre-specified to a limit of about 3.2 million people, the larger the sample size is set to, the greater the odds that the target population is represented with the data. In terms of weaknesses of the ACS dataset, there are none all that significant. Some minor concerns, however, are potential measurement error, processing error, and/or adjustment error in the data.

@@

```{r bills, fig.cap="Bills of penguins", echo = FALSE}

Talk more about it.

Also bills and their average (Figure \@ref(fig:billssssss)). (Notice how you can change the height and width so they don't take the whole page?)

```{r billssssss, fig.cap="More bills of penguins", echo = FALSE, fig.width=8, fig.height=4}

Talk way more about it.

@@

In order to forecast the popular vote of the U.S 2020 elections, multilevel logistic regression with post-stratification (MRP) was used. The individual survey gives sample data on voters of the general U.S population, and post-stratification allows us to reweight estimates, adjusting bias between our sample and the target population, so that we have a representative sample of likely voters @ForecastingArticle. To achieve this, cells are constructed using variables in the cleaned Nationscape survey such as age, household income, and race. The model is then trained on the survey using the proportions given by the  ACS post-stratification data set. This approach gives us an advantage when attempted to forecast voting, as we can use a broad survey to speak to subsets in the population, and also tends to be less expensive to collect than non-probability samples @rohansmrp. However, using this approach places limitations on how much we can interpret from the estimates. Although we can estimate voting intention in different demographic groups, it doesn't tell us any qualitative information on voting patterns. For example, we might be able to see how different age categories vote, but we won't have a measure of the policy preferences, or their views on the candidates.

Using the Nationscape,  a multilevel logistic regression model can be fitted to the survey data set to model the proportion of voters who will vote for Trump.
Logistic regression can be used to model a binary dependent variable, which is the choice between the Republic candidate, Donald Trump, and the Democratic candidate, Joe Biden in our case. This would be the main reason for choosing this model, over another model such as a linear regression model, where the output isn't necessarily binary. However, this imposes limitations. As we cannot take into consideration any other candidates running in the election due to the binary outcome. Another weakness is **** The variable 'vote_2020' in the data set was used as dependent variable, with the variable return 1 for voters intending to vote for Trump, and 0.

@@

We fit our regression model using the `glm()` function in [`R`].  We can see from table 1, which was made using `kable` from `knitr`,  that the fitted model results in the following coefficients:
```{r}
# loading data sets and cell counts for MRP
cell_counts <- readRDS(here::here("inputs/cleaned_data", "cell_counts.rds"))
individual_survey <- readRDS(here::here("inputs/cleaned_data","individual-survey.rds"))
post_strat <- readRDS(here::here("inputs/cleaned_data","post-strat.rds"))

@@

final_model <- readRDS(here::here("outputs/model","final_model.rds"))
coefs <- readRDS(here::here("outputs/model","coefficients.rds"))
knitr::kable(coefs, caption = " Table 1: Model Coefficients") %>% kableExtra::kable_styling(latex_options = "scale_down")

```
In logistic regression, collinearity implies that we have predictor variables that are highly correlated; that is, they have a linear relationship. This is possible when you have a larger number of variables in a model, but having collinearity in multiple variables can lead to unstable estimates and inaccurate variances, which can affect confidence intervals. This could lead to incorrect inferences about the relationship between our response variable and explanatory variables @doi:10.1080/09720502.2010.10700699. Thus, it is imperative to check for any correlated variables in our model to avoid these issues. The value of the Variance Inflation Factor(VIF) in for the explanatory variables can give us a measure of collinearity our model. VIF values should be less than 5 to guarantee that collinearity is not an issue. We can see in table 2 below that all VIF values are less than 5, so collinearity will not be a concern for our model. 
```{r}
cor <- readRDS(here::here("outputs/model","cor.rds"))
knitr::kable(cor, caption =  "Checking Model Collinearity", digits = 2)
```

```{r}
plot_model(final_model, vline.color = "red",  transform = NULL, title = "Figure _: Odds Ratios of the Logistic Regression model")
```

```{r}
# make predictions using our model with data from post-strat data set

@@

cell_counts$estimate <- predict(final_model, newdata = cell_counts, type = "response")

# post-stratified estimates - number of trump votes
estimate_votes <- cell_counts %>% 
  mutate(predict_prop = prop * estimate) %>% 
  group_by(state_name) %>% 
  summarise(predicted_vote = sum(predict_prop)) %>% ungroup()

estimate_votes

estimate_votes %>% mutate(status = if_else(predicted_vote>=0.50,"win","loss")) %>% group_by(status) %>% tally(
)

@@

*** add another diagnostic test

@@

```{r, echo=FALSE, include=FALSE}
# by state
estimates <- estimates %>% mutate(num_votes = n*estimate)
```

```{r age, echo=FALSE, include=FALSE, fig.cap="Distribution of Age between ACS and Nationscape data"}

individual_age_dist<- individual_survey %>% 
  group_by(age) %>% 
  summarise(num_records = n()) %>% 
  mutate(percentage = 100 * num_records/sum(num_records),
         total_num = sum(num_records),
         data_source = "Nationscape")

@@

```{r}
individual_hhincome_dist<- individual_survey %>% 
  group_by(household_income) %>% 
  summarise(num_records = n()) %>% 
  mutate(percentage = 100 * num_records/sum(num_records),
         total_num = sum(num_records),
         data_source = "Nationscape") %>% rename(hhincome_group = household_income)

post_strat_hhincome_dist <- post_strat %>% 
  group_by(hhincome_group) %>% 
  summarise(num_records = n()) %>% 
  mutate(percentage = 100 * num_records/sum(num_records),
         total_num = sum(num_records),
         data_source = "ACS")

individual_hhincome_dist %>% 
  union(post_strat_hhincome_dist) %>% 
  ggplot(aes(x = forcatshhincome_group, y = percentage)) + 
  geom_col() + 
  facet_grid(~data_source) + 
  theme_economist_white()

>>>>

library(tidyverse)
library(tinytex)
library(readr)
library(here)
library(dplyr)
library(ggplot2)
library(ggthemes)
library(ResourceSelection)
library(maps)
library(mapproj)
library(caret)
```

@@

.....

@@

 ## Individual Survey Data

 ## Post-stratification Data
As previously mentioned, our data is from two different datasets; the Democracy Fund + UCLA Nationscape Wave 50 dataset, and the 2018 1-year American Community Survey (ACS) dataset. They each contain both qualitative and quantitative data. To forecast the elections, we are using the Nationscape Wave 50 dataset. In regard to the Nationscape Wave 50 dataset, a few of its strengths are the variety in variables considered, and the quality of the survey responses. With respect to its weaknesses, it is likely that the data contains sampling error and/or coverage error. One of the various strengths of the  ACS dataset is its ability to be customized with pre-specified (or pre-selected?) variables and sample size before download. Although there is a wide range of variables to choose from, being able to select only those which we needed made the exploratory data analysis process much more manageable than it could have otherwise been (with all possible variables). Furthermore, since sample size can be pre-specified to a limit of about 3.2 million people, the larger the sample size is set to, the greater the odds that the target population is represented with the data. In terms of weaknesses of the ACS dataset, there are none all that significant. Some minor concerns, however, are potential measurement error, processing error, and/or adjustment error in the data.

@@

@@

In order to forecast the popular vote of the U.S 2020 elections, multilevel logistic regression with post-stratification (MRP) was used. MRP is useful for when generalizing from a possibly non-representative poll kennedy2020know.  The individual survey gives sample data on voters of the general U.S population, and post-stratification allows us to re-weight estimates, adjusting bias between our sample and the target population, so that we have a representative sample of likely voters @ForecastingArticle. To achieve this, cells are constructed using variables in the cleaned Nationscape survey such as age, household income, and race. The model is then trained on the survey using the proportions given by the  ACS post-stratification data set. This approach gives us an advantage when attempted to forecast voting, as we can use a broad survey to speak to subsets in the population, and also tends to be less expensive to collect than non-probability samples @rohansmrp. However, using this approach places limitations on how much we can interpret from the estimates. Although we can estimate voting intention in different demographic groups, it doesn't tell us any qualitative information on voting patterns. For example, we might be able to see how different age categories vote, but we won't have a measure of the policy preferences, or their views on the candidates.

Using the Nationscape data, a multilevel logistic regression model can be fitted to the survey data set to model the proportion of voters who will vote for Trump.
Logistic regression can be used to model a binary dependent variable, which is the choice between the Republic candidate, Donald Trump, and the Democratic candidate, Joe Biden in our case. This would be the main reason for choosing this model, over another model such as a linear regression model, where the output isn't necessarily binary. However, this imposes limitations. As we cannot take into consideration any other candidates running in the election due to the binary outcome. The variable 'vote_2020' in the data set was used as the dependent variable (the variable we are trying to predict), with the variable returning 1 for voters intending to vote for Trump, and 0 if they are voting for Biden.

@@

We fit our regression model using the `glm()` function in [`R`].  We can see from table 1, which was made using `kable` from `knitr`,  that the fitted model results in the following coefficients:
```{r, echo=FALSE}
# loading data sets and cell counts for MRP
cell_counts <- readRDS(here::here("inputs/cleaned_data", "cell_counts.rds"))
individual_survey <- readRDS(here::here("inputs/cleaned_data","individual-survey.rds"))
post_strat <- readRDS(here::here("inputs/cleaned_data","post-strat.rds"))

@@

final_model <- readRDS(here::here("outputs/model","final_model.rds"))
coefs <- readRDS(here::here("outputs/model","coefficients.rds"))
knitr::kable(coefs, caption = " Table 1: Model Coefficients", digits = 2)

```
In logistic regression, collinearity implies that we have predictor variables that are highly correlated; that is, they have a linear relationship. This is possible when you have a larger number of variables in a model, but having collinearity in multiple variables can lead to unstable estimates and inaccurate variances, which can affect confidence intervals. This could lead to incorrect inferences about the relationship between our response variable and explanatory variables @doi:10.1080/09720502.2010.10700699. Thus, it is imperative to check for any correlated variables in our model to avoid these issues. The value of the Variance Inflation Factor(VIF) in for the explanatory variables can give us a measure of collinearity our model. VIF values should be less than 5 to guarantee that collinearity is not an issue. We can see in table 2 below that all VIF values are less than 5, so collinearity will not be a concern for our model. 
```{r, echo=FALSE}
cor <- readRDS(here::here("outputs/model","cor.rds"))
knitr::kable(cor, caption =  "Checking Model Collinearity", digits = 2)
```

```{r, echo=FALSE}
 htst <- hoslem.test(final_model$y, fitted(final_model), g=10)
htst
```
Before the model can be used for any predictions, we must also check that our model fits our data well, and meets any assumptions made by the model. The Hosmer-Lemeshow goodness of fit test assesses whether or not the observed events match expected events in subgroups of the model. For this test, the p-value can be used to assess goodness of fit. A smaller p-value indicates strong evidence against a good fit. In the case of our model, we can see in table 4 that we have a very large p-value of 0.9, indicating no evidence of poor fit.

| Table 4: Hosmer and Lemeshow goodness of fit (GOF) test 	|    	|         	|
|---------------------------------------------------------	|----	|---------	|
| X-squared                                               	| df 	| p-value 	|
| 3.5                                                     	| 8  	| 0.9     	|

```{r, echo=FALSE}
# make predictions using our model with data from post-strat data set

@@

cell_counts$estimate <- predict(final_model, newdata = cell_counts, type = "response")
#  prediction intervals
# fitted value - error to get margin of error
se_md <- predict(final_model, newdata = cell_counts, type = "response", se.fit = T)

 lwr <-  cell_counts$estimate - 1.96*se_md$se.fit
 upper <- cell_counts$estimate + 1.96*se_md$se.fit

# all estimates with lower and upper bounds
estimates_with_margins <- cell_counts %>% mutate(n = num_records * estimate) %>% cbind(cell_counts, lwr, upper)
# removing duplicate column name estimate
estimates_with_margins <- estimates_with_margins[, !duplicated(colnames(estimates_with_margins))]

# post-stratified estimates - number of trump votes with prediction intervals
post_strat_estimates <- estimates_with_margins %>% 
  mutate(predict_prop = prop * estimate, l_predict = lwr * prop, u_predict = upper * prop) %>% 
  group_by(state_name) %>% 
  summarise(predicted_vote = sum(predict_prop), lower = sum(l_predict), upp = sum(u_predict)) %>% group_by(state_name) %>% ungroup()

post_strat_estimates %>% mutate(status = if_else(predicted_vote>=0.50,"win","loss")) %>% group_by(status) %>% tally(
)

@@

@@

```{r age, echo=FALSE, include=FALSE, fig.cap="Distribution of Age between ACS and Nationscape data"}
individual_age_dist<- individual_survey %>% 
  group_by(age) %>% 
  summarise(num_records = n()) %>% 
  mutate(percentage = 100 * num_records/sum(num_records),
         total_num = sum(num_records),
         data_source = "Nationscape")

@@

```{r, echo = FALSE}

# estimated proportion of votes by State
post_strat_estimates %>%
ggplot(aes(y = predicted_vote, x =state_name, color = "MRP estimate")) + geom_point() +
geom_errorbar(aes(ymin = lower, ymax = upp), width = 0) +
ylab("Proportion of Trump Supporters") +
xlab("State") +
geom_point(data = individual_survey %>%
group_by(state_name, vote_2020) %>% summarise(n = n()) %>% group_by(state_name) %>%
mutate(prop = n/sum(n)) %>% filter(vote_2020==1),
aes(state_name, prop, color = "Raw data")) + theme_minimal() +
scale_color_brewer(palette = "Set1") + theme(legend.position = "bottom") + theme(legend.title = element_blank()) + 
  theme(axis.text.x = element_text(angle = 35, hjust = 1, size = 5)) + coord_flip() +labs(title = "2020 Predicted Popular Vote by State", subtitle = "Compating redicted popular vote by state using raw data and MRP")
```

```{r, echo = FALSE}
# Sorting State wins
election_wins <- post_strat_estimates %>% mutate(winner = if_else(predicted_vote>=0.50,"Trump","Biden")) %>% group_by(state_name
)
election_wins$state <- str_to_title(as.character(election_wins$state_name))
election_wins <- election_wins %>% filter(state != "district of columbia")

# Plotting state wins
statebins(election_wins, value_col="winner", ggplot2_scale_function = scale_fill_manual, values = c(Trump = "#2166ac", Biden = "#b2182b"), name = "state") +
  theme_statebins(legend_position="right")
